Power simulation section

### 2023-09-06 docs/manuscript.Rmd
further improved manuscript
went through manuscript again and made final edits before handing over to Flo

---
title             : "A Tutorial for Customized Power Simulations for Complex Experimental Study Designs using Generalized Linear Mixed Models"
shorttitle        : "GLMM Power Simulation Tutorial"

author: 
  - name          : "Timo K. Koch"
    affiliation   : "1, 2, 3"
    corresponding : yes    # Define only one corresponding author
    #address       : "Postal address"
    email         : "timo.koch@unisg.ch"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Formal Analysis"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Florian Pargent"
    affiliation   : "2"
    role:
      - "Formal Analysis"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Anne-Kathrin Kleine"
    affiliation   : "1"
    role:
      - "Writing - Review & Editing"
  - name          : "Susanne Gaube"
    affiliation   : "1, 4"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "LMU Munich, Center for Leadership and People Management"
  - id            : "2"
    institution   : "LMU Munich, Department of Psychology"
  - id            : "3"
    institution   : "University of St. Gallen, Institute of Behavioral Science & Technology"
  - id            : "4"
    institution   : "University College London, Global Business School for Health"

@@

  Understanding how humans interact with artificial intelligence (AI) through experiments becomes increasingly relevant. However, experimental interaction researchers often lack the appropriate tools to conduct power and data analyses for the complex study designs required. In this work, we provide a comprehensive guide to performing customized power analyses using data simulation based on Generalized Linear Mixed Models (GLMMs). By providing code in a specific case study, we equip experimental interaction researchers with the tools to simulate their own data and run analyses based on GLMMs. We discuss implications for experimental interaction research and provide an outlook on the relevance of power simulations in the field.

@@

wordcount         : "6551"

@@

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research it is essential to determine an appropriate sample size [@lakensSampleSizeJustification2022] to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data. However, the recent replication crisis in experimental research, such as in Human-Computer Interaction (HCI), has illustrated many challenges surrounding the reproducibility and reliability of findings. As a results, there is a growing need for rigorous statistical methodology and the adoption of well-powered experimental designs. While accessible software solutions exists for simple statistical models and experimental designs, many experimental researcher lack the skills and tools to conduct power analyses for more complex research endeavors, such as powerful generalized linear mixed models (GLMM). In the present case study, we provide a tutorial comprised of a concrete example for custom power simulations using GLMMs.

@@

## Power analysis

<!-- define power analysis -->
A power analysis represents the act of calculating the statistical power for a given true effect and sample size. By running the calculation for a range of sample sizes, the required sample size can be determined so that researchers to find an assumed true effect with the desired statistical power. 
Power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes required to achieve sufficient statistical power for results with high evidential value. 
Moreover, conducting a careful a priori power analysis helps researchers decide which experimental design and statistical models are both feasible and appropriate for answering their research questions and analysing the data.

<!-- advantages of an a priori power analysis -->
For a desired statistical power and an assumed true effect, the required sample size to find such an effect can be calculated a priori. Thereby, power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes needed to achieve sufficient statistical power for reliable results. Also, while conducting a proper power analysis, one has to consider every aspect of the experimental design. This really makes one think through everything carefully and might highlight statistical or design challenges before the start of an experiment. This gives the researcher a great opportunity to improve the design. Especially in fast-moving research fields, where speed is often regarded as more important than rigor, adding a proper sample size calculation into the process can act as a safeguard for ensuring high-quality research.

@@

<!-- Advantages of GLMMs over "simpler" models-->
However, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. Therefore, GLMMs (generalized) linear mixed models (GLMMs; also called multilevel models) are gaining increasing popularity in analyzing data from experimental research, such as in HCI. GLMMs offer a flexible framework for analyzing data with non-normal and categorical outcomes, accounting for both fixed and random effects.

<!-- intro into GLMMs -->
LMMs (Linear Mixed Models) and GLMMs (Generalized Linear Mixed Models) are powerful statistical frameworks that handle complex data structures by incorporating both fixed and random effects [@fahrmeirRegressionModelsMethods2021]. LMMs are extensions of linear regression models that account for correlated data and hierarchical structures. In this context, correlated data means that the observations within a given dataset are not independent of each other. In other words, the values of the dependent variable for one observation may be related to the values of the dependent variable for other observations in a systematic way. This correlation can arise for various reasons, such as repeated measurements on the same subjects over time, observations clustered within certain groups or locations, or data collected at different levels of granularity. Hierarchical structures can be found in datasets that have a nested or hierarchical organization. For example, one might have data on students nested within schools, employees nested within departments, or patients nested within hospitals. In such cases, the observations within each higher-level grouping (e.g., school, department, hospital) tend to be more similar to each other than to observations in other groups.
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on the predictor variables). They allow for the modelling of fixed effects, which capture the relationships between predictors and the outcome, as well as random effects, which account for the correlation and variability within groups or subjects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across the groups or subjects. GLMMs extend the LMM framework to accommodate non-normal and categorical outcome variables.  They are used when the outcome variable does not follow a normal distribution, but instead belongs to a different distribution family (e.g., binomial). GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear predictor to the expected value of the outcome variable. The link function allows for modelling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the response variable.

@@

<!-- what are the existing option for a priori power analysis for GLMMs / where are existing tutorials limited -->
Due to the previously mentioned advantages of GLMMs, researchers are increasingly using those for their statistical analyses. 
However, available software packages for power simulations do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022]. 
As a consequence, custom power simulations are needed.

A number of tutorials have been published describing how to perform power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE), which are very common [@brownIntroductionLinearMixedEffects2021]. Moreover, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, rendering estimation a practical challenge [@maxwellSampleSizePlanning2008].

<!-- basics of power simulation for GLMMs  -->
Power simulations in GLMMs are essential for estimating the statistical power of complex experimental study designs. The model equation for a GLMM combines fixed effects, random effects, and an appropriate link function to model the relationship between predictors and the outcome variable. The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed.
When planning an experiment with complexities such as the use of GLMMs, non-standard designs (e.g., hierarchical structure), and "missing by design" conditions, the process of power simulations becomes challenging. 
Unfortunately, there are no general recommendations that can be applied universally in such cases. Instead, researchers must acquire the necessary skills based on the specific application at hand.
By incorporating GLMMs and power analysis into research planning, researchers can enhance the replicability and credibility of their work, [@yarkoniGeneralizabilityCrisis2022].

<!-- Summary / Wrap-Up, pick up main arguments and what the paper provides, keep this short -->
Based on need for sufficiently powered experimental research and the lack of tools to run custom power analyses for complex study designs using GLMMs, in this paper, we present a case study that serves as a practical demonstration of how to perform a data simulation for a concrete study. Through this case study, we aim to provide guidance and insights into the complexities involved in addressing sophisticated design considerations and obtaining accurate required sample size estimates.

# Method

@@

In this section, we outline the steps for performing data simulation and power analysis for GLMMs, using a case study based on a specific experimental study design from the area of experimental interaction research.

@@

<!-- describe the study idea and design -->
In the case study, we simulate the data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system is to be evaluated. 
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts), review head CT (computer tomography) scans to assess the presence of bleeding, more specifically, intracranial haemorrhage (ICH). 
To support their decision-making, in two experimental conditions, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate. Each participant undergoes all experimental conditions, separated by a two-week washout period to prevent scan recall.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans, (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice, and (c) different presentations of AI advice influence performance.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the power simulation is to determine how many task experts and non-experts to recruit to achieve sufficient statistical power in our experiment.

@@

In our case study, the probability of giving a correct response is assumed to depend on the predictors: <!-- SG: What happened to the different experimental conditions? AI advice had three levels: Advice with annotation, without annotation, and no advice--> - $advice\_present_{si}$: whether subject $s$ was presented with AI advice (1) or not (0) when asked to asses item $i$ - $advice\_correct_{si}$: whether this advice was correct (1) or not (0) - $expert_s$: whether subject $s$ was a task expert (1) or not (0)

@@

<!-- SG: You basically start the explanation at line 257; I think it would be better to actually start at the beginning of the function line 251, otherwise, this might be very confusing to people. If you do not want to change the structure, then I would at least mention that you discuss the first section of the function below.-->

@@

When introducing our simulation function and simulating data for the above example, we have used theoretically plausible values as defaults for all model parameters ($\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$), but have not talked about where these numbers came from. All parameter values have been determined in repeated exchanges with our affiliated domain experts, and we now outline a few strategies on how to determine plausible parameter values.

We have already seen in our discussion of model interpretation how we can derive the model implied probability for each experimental condition, that a subject with average ability solves an item with average difficulty. We can revert this perspective by choosing plausible probability values and deriving the parameter values implied by these probabilities (for an average subject and an average item).

Often no meta-analytic results or conclusive data from pilot studies are available. Then domain knowledge is essential.

@@

Although the derivations above are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty. For example, we determined the $\beta$ parameters in a way that corresponds to a desired probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice. c However, even if the model were true, we would not observe this probability if we estimated the probability in a group of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty. This implies that one must be careful when specifying parameter values based on previous studies or pilot data. <!-- SG: What does that mean for someone who wants to do this then?-->

The inequality <!-- SG: Can you explain what this means?--> of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult. However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge. Then we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

Power analysis can quickly become computationally intensive when we repeatedly simulate data and fit models for different parameter combinations or sample sizes. Thus, we use the future [@R-RJ-2021-048] and furrr [@R-furrr] packages to perform computations in parallel. First, we enable parallelization and specify how many parallel cores of our computer to use (users can find out the maximum number of cores on their computer with the command `parallel::detectCores()`), and set a seed to make the simulation reproducible. <!-- SG: Are workers the cores you want to use?-->

```{r, message=FALSE, eval=FALSE}
library(future)
plan("multisession", workers = 6)
set.seed(2)
```

The next code chunk specifies a simulation grid with different settings for both the number of subjects (`n_subjects`) and the number of items (`n_items`), each combination being repeated `rep` times. <!-- SG: Can you explain how you choose the repetitions?-->

@@

Experimental research requires careful planning and consideration of statistical power to ensure reliable and meaningful results. While heuristics and user-friendly software can be useful for simple designs and models, they often fall short when more complex and customized simulations are required. The present tutorial presents a specific example of how to run power simulations for experimental interaction experiments, which can also applied to other fields of research.

@@

1.  The need for power simulations in experimental research: Experimental research often involves intricate designs and complex models that cannot be adequately addressed by heuristics or user-friendly software. Power simulations offer a solution by providing experimental interaction researchers with a tailored approach to estimating statistical power. These simulations take into account the specific study design, account for the underlying assumptions, and offer more accurate power estimates.

@@

3.  Teaching power simulation skills: The ability to conduct power simulations is a valuable skill that should be taught to experimental interaction researchers. By incorporating this training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power and improve the quality of their experimental designs. 
Equipping researchers with the knowledge and tools to perform power simulations enables them to make informed decisions and enhance the rigour of their studies. 
The need to reason about how to simulate plausible data that is in line with the research hypothesis while not violating domain expertise on how plausible data should look might also contribute to planning more insightful studies that can answer more precise research questions [@yarkoniGeneralizabilityCrisis2022].

4.  Addressing the mismatch in effort perception: There is often a significant disconnect between the amount of effort required to perform power analysis and the perceived effort estimated by researchers and collaborators in experimental interaction research. Many researchers request power simulations from statisticians or methodological experts without fully comprehending the complexity and time-consuming nature of these tailored simulations. It is crucial to raise awareness about the effort involved in power simulations to ensure realistic expectations and effective collaboration between researchers and methodological experts.

5.  Recognizing the value of power simulations: Power simulations are not mere technicalities; they are valuable research contributions that deserve recognition in experimental interaction research. They offer insights into the robustness and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power. The importance of power simulations can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports.

6.  Integration with Open Science and preregistration practices: Power simulations align well with the principles of Open Science and preregistration in experimental interaction research. When researchers have access to simulated data based on their prespecified model, analyzing the collected dataset becomes straightforward and unambiguous. By preregistering their power simulations, researchers enhance transparency and accountability in their experimental procedures, contributing to the credibility and reproducibility of research.

@@

Power simulations play a critical role in experimental research, allowing investigators to tailor power estimation to the unique aspects of their experiments. Through this tutorial, we aim to provide researchers with the necessary skills and tools to perform these simulations themselves. By integrating power simulations with open science and preregistration practices, researchers can improve the robustness and transparency of their findings, advancing the field.

>>>>

---
title             : "A Tutorial on Tailored Power Simulations for Experimental Study Designs with Generalized Linear Mixed Models"
shorttitle        : "GLMM Power Simulation Tutorial"

author: 
  - name          : "Timo K. Koch"
    affiliation   : "1, 2"
    corresponding : yes    # Define only one corresponding author
    #address       : "Postal address"
    email         : "timo.koch@unisg.ch"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Formal Analysis"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Florian Pargent"
    affiliation   : "1"
    role:
      - "Formal Analysis"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Anne-Kathrin Kleine"
    affiliation   : "1"
    role:
      - "Validation"
      - "Writing - Review & Editing"
  - name          : "Eva Lermer"
    affiliation   : "1,3"
    role:
      - "Resources"
  - name          : "Susanne Gaube"
    affiliation   : "1,4"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "LMU Munich, Department of Psychology"
  - id            : "2"
    institution   : "University of St. Gallen, Institute of Behavioral Science & Technology"
  - id            : "3"
    institution   : "Technical University of Applied Sciences Augsburg, Department of Business Psychology"
  - id            : "4"
    institution   : "University College London, Global Business School for Health"

@@

  [THIS WILL BE UPDATED BASED ON GOOGLE DOC] Understanding how humans interact with artificial intelligence (AI) through experiments becomes increasingly relevant. However, experimental interaction researchers often lack the appropriate tools to conduct power and data analyses for the complex study designs required. In this work, we provide a comprehensive guide to performing customized power analyses using data simulation based on Generalized Linear Mixed Models (GLMMs). By providing code in a specific case study, we equip experimental interaction researchers with the tools to simulate their own data and run analyses based on GLMMs. We discuss implications for experimental interaction research and provide an outlook on the relevance of power simulations in the field.

@@

wordcount         : "7059"

@@

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research it is essential to determine an appropriate sample size [@lakensSampleSizeJustification2022] to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data. However, the recent replication crisis in experimental research, such as in Human-Computer Interaction (HCI), has illustrated many challenges surrounding the reproducibility and reliability of findings. As a results, there is a growing need for rigorous statistical methodology and the adoption of well-powered experimental designs. While accessible software solutions exists for simple statistical models and experimental designs, many experimental researcher lack the skills and tools to conduct power analyses for more complex research designs using powerful generalized linear mixed models (GLMM). In the present case study, we provide a tutorial comprised of a concrete example for custom a priori power analyses using data simulations based on GLMMs.

@@

## A priori power analysis

<!-- define power analysis -->
A power analysis represents the act of calculating the statistical power for a given true effect and sample size. 
By running the calculation for a range of sample sizes "a priori" (i.e., beforehand), the required sample size can be determined so that researchers find an assumed true effect with the desired statistical power. 
Thereby, power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes required to achieve sufficient statistical power for results with high evidential value.

<!-- advantages of an a priori power analysis -->
Conducting a careful a priori power analysis helps researchers decide which experimental design and statistical models are both feasible and appropriate for answering their research questions and analyzing the data.
Also, while conducting a proper power analysis, one has to consider every aspect of the experimental design. This makes one think through everything carefully and might highlight statistical or design challenges before the start of an experiment. 
This gives researchers a great opportunity to improve the design. Especially in fast-moving research fields, where speed is often regarded as more important than rigor, adding a solid sample size calculation into the process can act as a safeguard for ensuring high-quality research.

@@

<!-- Advantages of GLMMs over "simpler" models -->
As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. As a result, GLMMs (generalized) linear mixed models (GLMMs; also called multilevel models) are gaining increasing popularity in analyzing data from experimental research, such as in HCI, because they offer a flexible framework for analyzing data with non-normal and categorical outcomes while accounting for both fixed and random effects [@kapteinUsingGeneralizedLinear2016; @fahrmeirRegressionModelsMethods2021].

<!-- intro into GLMMs -->
GLMMs are an extension of LMMs (Linear Mixed Models) which are in turn extensions of classic linear regression models that account for correlated data and hierarchical structures [@fahrmeirRegressionModelsMethods2021]. In this context, correlated data means that observations within a given dataset are not independent of each other. In other words, the values of the dependent variable for one observation may be related to the values of the dependent variable for other observations in a systematic way. This correlation can arise for various reasons, such as repeated measurements on the same subjects over time, observations clustered within certain groups or locations, or data collected at different levels of granularity. Hierarchical structures can be found in datasets that have a nested or hierarchical organization. For example, one might have data on students nested within schools, employees nested within departments, or patients nested within hospitals. In such cases, the observations within each higher-level grouping (e.g., school, department, hospital) tend to be more similar to each other than to observations in other groups.
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on the predictor variables). They allow for the modelling of fixed effects, which capture the relationships between predictors and the outcome, as well as random effects, which account for the correlation and variability within groups or subjects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across the groups or subjects. 
As mentioned, GLMMs extend the LMM framework to accommodate non-normal and categorical outcome variables. They are used when the outcome variable does not follow a normal distribution, but instead belongs to a different distribution family (e.g., binomial). GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear predictor to the expected value of the outcome variable. The link function allows for modelling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the response variable.

@@

<!-- what are the existing option for a priori power analysis for GLMMs / where are existing tutorials limited -->
Currently available software packages for power simulations do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022]. 
A number of tutorials have been published describing how to perform such power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE), which are very common [@brownIntroductionLinearMixedEffects2021]. 
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Here, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, rendering estimation a practical challenge [@maxwellSampleSizePlanning2008]. 
<!-- Summary / Wrap-Up, pick up main arguments and what the paper provides, keep this short! -->
Based on need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a data simulation for a priori power simulations of such study designs. Thereby, we aim to equip researchers with the tools needed to simulate their own data and determine appropriate sample sizes with GLMMs.

# The present case study

@@

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs, using a case study based on a specific experimental study design from the area of human-AI interaction research.

@@

<!-- describe the study idea and design -->
In the present case study, we simulate the data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system is to be evaluated. 
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts), review head CT (computer tomography) scans to assess the presence of bleeding, more specifically, intracranial haemorrhage (ICH). 
To support their decision-making, in two experimental conditions, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate. Each participant undergoes all experimental conditions.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans, (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice, and (c) different presentations of AI advice influence performance.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the power simulation is to determine how many task experts and non-experts to recruit to achieve sufficient statistical power in our experiment.

@@

In our case study, the probability of giving a correct response is assumed to depend on the predictors:

- $advice\_present_{si}$: whether subject $s$ was presented with AI advice (1) or not (0) when asked to asses item $i$ 
- $advice\_correct_{si}$: whether this advice was correct (1) or not (0) 
- $expert_s$: whether subject $s$ was a task expert (1) or not (0)

@@

@@

When introducing our simulation function and simulating data for the above example, we have used theoretically plausible values as defaults for all model parameters ($\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$), but have not talked about where these numbers came from.

Often no meta-analytic results or conclusive data from pilot studies are available. All parameter values in our present case study have been determined based on results from prior own work and from the literature. Additionally, we had repeated exchanges with our affiliated domain experts from radiology.

We now outline a few strategies on how to determine plausible parameter values. We have already seen in our discussion of model interpretation how we can derive the model implied probability for each experimental condition, that a subject with average ability solves an item with average difficulty. We can revert this perspective by choosing plausible probability values and deriving the parameter values implied by these probabilities (for an average subject and an average item).

@@

Although the derivations above are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty. For example, we determined the $\beta$ parameters in a way that corresponds to a desired probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice. However, even if the model were true, we would not observe this probability if we estimated the probability in a group of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty. This implies that one must be careful when specifying parameter values based on previous studies or pilot data. <!-- SG: What does that mean for someone who wants to do this then?-->

The inequality of conditional and marginal effects in GLMMs (i.e., how predictor variables affect the response within groups versus their overall impact across all groups, which can differ due to random effects) [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult. However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge. Then we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

Power analysis can quickly become computationally intensive when we repeatedly simulate data and fit models for different parameter combinations or sample sizes. Thus, we use the future [@R-RJ-2021-048] and furrr [@R-furrr] packages to perform computations in parallel. First, we enable parallelization and specify how many parallel cores ("workers") of our computer to use (users can find out the maximum number of cores on their computer with the command `parallel::detectCores()`), and set a seed to make the simulation reproducible.

```{r, message=FALSE, eval=FALSE}
library(future)
plan("multisession", workers = 4)
set.seed(2)
```

The next code chunk specifies a simulation grid with different settings for both the number of subjects (`n_subjects`) and the number of items (`n_items`), each combination being repeated `rep` times. We chose 300 repetitions for the data simulation at hand as it strikes a balance between achieving a robust statistical estimate and remaining computationally feasible.

@@

Experimental research requires careful planning and consideration of statistical power to ensure reliable and meaningful results. While heuristics and user-friendly software can be useful for simple designs and models, they often fall short when more complex and customized simulations are required. The present tutorial presents a specific example of how to run power simulations for experimental designs with GLMMs.

@@

1.  The need for power simulations in experimental research: Experimental research often involves intricate designs and complex models that cannot be adequately addressed by heuristics or user-friendly software. Power simulations offer a solution by providing experimental researchers with a tailored approach to estimating statistical power. These simulations take into account the specific study design, account for the underlying assumptions, and offer more accurate power estimates.

@@

3.  Teaching power simulation skills: The ability to conduct power simulations is a valuable skill that should be taught to experimental researchers. By incorporating this training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power and improve the quality of their experimental designs. 
Equipping researchers with the knowledge and tools to perform power simulations enables them to make informed decisions and enhance the rigour of their studies. 
The need to reason about how to simulate plausible data that is in line with the research hypothesis while not violating domain expertise on how plausible data should look might also contribute to planning more insightful studies that can answer more precise research questions [@yarkoniGeneralizabilityCrisis2022].

4.  Addressing the mismatch in effort perception: There is often a significant disconnect between the amount of effort required to perform power analysis and the perceived effort estimated by researchers and collaborators in experimental research. Many researchers request power simulations from statisticians or methodological experts without fully comprehending the complexity and time-consuming nature of these tailored simulations. It is crucial to raise awareness about the effort involved in power simulations to ensure realistic expectations and effective collaboration between researchers and methodological experts.

5.  Recognizing the value of power simulations: Power simulations are not mere technicalities; they are valuable research contributions that deserve recognition in experimental research. They offer insights into the robustness and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power. The importance of power simulations can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports.

6.  Integration with Open Science and preregistration practices: Power simulations align well with the principles of Open Science and preregistration in experimental research. When researchers have access to simulated data based on their prespecified model, analyzing the collected dataset becomes straightforward and unambiguous. By preregistering their power simulations, researchers enhance transparency and accountability in their experimental procedures, contributing to the credibility and reproducibility of research.

@@

Power simulations play a critical role in experimental research, allowing investigators to tailor power estimation to the unique aspects of their experiments. Through this tutorial, we aim to provide researchers with the necessary skills and tools to perform these simulations for GLMMs themselves. By incorporating GLMMs and power analysis into research planning, researchers can enhance the replicability and credibility of their work [@yarkoniGeneralizabilityCrisis2022].
